---
title: "Quiz 7: CLV - Non-contractual settings"
author: "Daniel Redel"
date: "2023-01-26"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("scipen"=100, "digits"=3, width = 300)
rm(list=ls())
library(readr)
library("BTYD")
```

In this assignment, we will use a different data set with 10000 customers and [*13 periods total*]{.underline}. We estimate the model on the **first 7 (calibration)** and test it on the **final 6 periods (holdout)**.  

```{r}
cal.rf.matrix <- read.csv("cal.rf.matrix.csv")
trans <- read.csv("annual_transactions.csv")
holdout <- read.csv("holdout.trans.rf.matrix.csv")
```

# Question 1

Create a graph of aggregate transactions per year. **What are the total number of purchases in year 8 after the first purchase?**

*Provide your answer without any decimals (e.g. 12)*

```{r}
trans <- trans$x

par(mfrow=c(1,1))
par(mai=c(.8,.8,.2,.2))
plot(seq(1,13,1),trans, type="b", ylab="Total number of repeat transactions", xlab="Year since first purchase", main="", xaxt='n')
axis(1, at = seq(0, 13, by = 1))
abline(v=7.5,col = "red", lwd = 2)
text(x = 6,y = 4500,"Calibration", cex=1, pos=3, col="black", font = 2)
text(x = 9,y = 4500,"Validation", cex=1, pos=3, col="black", font = 2)

```

# Question 2

Estimate the BG/BB parameters using the calibration data, the first seven years. **What are the parameters?**

*Provide your answers with a dot and two decimals (e.g. 0.12)*

**Initial Parameters**:

```{r}
rf.matrix <- cal.rf.matrix
par.start <- c(1, 1, 1, 1)

```

**MLE Estimation**:

```{r}
params <- bgbb.EstimateParameters(rf.matrix, par.start)
round(params, 2)
```

```{r}
## Check log-likelihood of the params:
LL <- bgbb.rf.matrix.LL(params, rf.matrix)
LL
```

# Question 3

Graph out the distributions of the transaction rate and dropout rate. **What are the average values of each in the population?**

*Provide your answer with a dot and two decimals (e.g. 0.12)*

[**Average transaction rate**]{.underline}:

```{r}
round(params[1]/(params[1]+params[2]),2)
bgbb.PlotTransactionRateHeterogeneity(params)
```

[**Average dropout rate**]{.underline}:

```{r}
round(params[3]/(params[3]+params[4]),2)
bgbb.PlotDropoutRateHeterogeneity(params)
```

# Question 4

Graph out the actual and predicted transactions per year. **What is the expected level of transactions in year 8 after the first purchase?**

*Provide your answer with zero decimals (e.g. 12)*

[**Aggregate Forecasting**]{.underline}:

```{r}
pred <- bgbb.PlotTrackingInc(params, rf.matrix, trans)

```

```{r, echo=FALSE}
cat("Q4 =",round(pred[2,8]), "\n")
```

# Question 5

According to the model, **what is the predicted number of transactions in the holdout for a customer who was "7 for 7" in the calibration period?  What is the actual number of holdout transactions for these customers?**

*Provide your answer with a dot and two decimals (e.g. 0.12)*

[**Predicted**]{.underline}:

```{r}
bgbb.HeatmapHoldoutExpectedTrans(params, n.cal = 7, n.star = 6)
```

[**Actual**]{.underline}:

```{r}
cbind(cal.rf.matrix, holdout)

# take total transactions for last row divided by customers:

act <- holdout$x[29]/cal.rf.matrix[29,4] #ind=29 has 7 of 7, the last obs.

```

```{r, echo=FALSE}
cat("Actual =",round(act,2), "\n")
```

# Question 6

**What is the CLV**, assuming each transaction yields on average 50 in profit and the discount rate is 0.1? You can use 200 periods for the sum.

*Provide your answer with zero decimals (e.g. 12)*

```{r}
BGBBCLV<-function(params,m,d,T) {
params<-unname(params)
al<-params[1]
be<-params[2]
ga<-params[3]
de<-params[4]
DET<-1   # at time zero there has to be a purchase
for (i in 1:T) {
    DET<-DET+(al/(al+be))*(beta(ga,de+i)/beta(ga,de))*1/(1+d)^{i}
}
CLV=m*DET  # convert discount expected purchases into expected value
return(CLV)    #return the CLV
}
```

[**CLV**]{.underline}:

```{r}
CLV <- BGBBCLV(params = params, m=50,d=.1,T=200)
CLV
```

# Question 7

**What is the RLV for a 7 for 7 customer?** Again, assume each transaction yields on average 50 in profit, and the discount rate is 0.1.  

*Provide your answer with zero decimals (e.g. 12)*

```{r}
m <- 50
DERT <- bgbb.rf.matrix.DERT(params, rf.matrix = cal.rf.matrix, d=0.1)
RLV <- m*DERT

RLVmatrix <- cbind(cal.rf.matrix,round(RLV)) 
RLVmatrix
```

[**RLV**]{.underline}:

```{r}
max(RLV)
```

# Question 8

Consider a charity with just one opportunity for people to donate per year. Consider a donor has donated 3 times in the past five years with the last donation occurring in year 4. In the context of the BG/BB model, what are the donor's sufficient statistics: n, t_x, and x

n=5\
t_x = 4\
x = 3

Assume that = 0.6 and *Θ* = 0.2.

**What is the likelihood of this customer's donations?**

*Provide your answer with a dot and three decimals (e.g. 0.123)*

```{r}
n <- 5
t_x <- 4
x <- 3

p <- 0.6
theta <- 0.2
```

$$
L(p,\theta|x, t_x,n)=p^x(1-p)^{n-x}(1-\theta)^n
+\sum^{n-t_x-1}_{j=0}p^x(1-p)^{t_x-x+j}\theta(1-\theta)^{t_x+j}
$$

```{r}
a <- (p^3)*(1-p)*((1-theta)^4)*theta
b <- (p^3)*(1-p)^2*((1-theta)^5) ## the last obs. does not have theta

likelihood <- a+b
round(likelihood,3)
```

# Question 9

**What is the probability that the customer is alive at the end?**

*Provide your answer with a dot and three decimals (e.g. 0.123)*

$$
\Pr(\text{alive at }n|x,t_x,n)=\frac{p^x(1-p)^{n-x}(1-\theta)^n}{L(p,\theta|x,t_x,n)}
$$

```{r}
(p^3)*((1-p)^2)*((1-theta)^5)/likelihood
```
