---
title: "Quiz 1: Test & Roll"
author: "Daniel Redel"
date: "2023-01-24"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
# PACKAGES
library(tidyverse)
library(kableExtra)
library(readr)
library(haven)
```

# Question 1

A company is considering whether to do a test on its marketing campaign. Direct marketing costs €2.00 to send. The company sells a product for €50 and 34% of that remains after taking costs and other items into account.

**What's the breakeven response rate threshold, i.e., the response probability this company needs for its mailing to be profitable?**

$$
\pi\hat{p}-c>0
$$

*Provide your answer with three decimals separated by a dot, not a comma (e.g. 0.123).*

```{r}
m <- 50*0.34
c <- 2
brk <- c/m
```

```{r, echo=FALSE}
cat("Breakeven Response Rate Threshold:", round(brk,3) )
```

# Question 2

A company is considering whether to do a test on its marketing campaign. Direct marketing costs €2.00 to send. The company sells a product for €50 and 34% of that remains after taking costs and other items into account.

**Using the *`ebeer`* data set, estimate the response rate.**

*Provide your answer with three decimals separated by a dot, not a comma (e.g. 0.123).*

The data:

```{r, warning=FALSE, message=FALSE}
ebeer <- read_sav("C:/Users/danny/OneDrive/Análisis Cuantitativo y Econometría/Marketing Analytics/Customer Analytics/Quizes/1 Uncertainty/E-Beer.sav")

# Eliminate NA's
ebeer <- ebeer[!is.na(ebeer$respmail),]
# Response is numeric and 0/1
ebeer$respmail <- as.numeric(ebeer$respmail)
```

**Response Rate**:

```{r}
p_hat <- round(mean(ebeer$respmail, na.rm=TRUE),3)
p_hat_se <- sqrt(p_hat*(1-p_hat)/nrow(ebeer))
```

```{r, echo=FALSE}
cat("The mean response rate is", round(p_hat,3), "and it's standard deviation is", round(p_hat_se, 3))
```

Assuming a *normal distribution*, we can estimate the probability that our estimated response rate is below the breakeven point:

```{r}
B <- 10000
set.seed(19103)
post_draws <- rnorm(B, mean = p_hat, sd = p_hat_se) # random deviates
prob <- sum(post_draws<brk)/B
```

```{r, echo=FALSE}
cat("The probability of being below the breakeven point is", round(prob, 3))
```

We can plot these results:

```{r}
xx <- seq(.107,.14,length=1000)

par(mai=c(.9,.8,.2,.2))
plot(xx, dnorm(xx, p_hat, p_hat_se), main="Distribution of Sample Mean", 
     type="l", col="royalblue", lwd=1.5,
     xlab="average monetary value", ylab="density")
abline(v=brk)
text(x = .115,y= 68, paste("P(p < 0.118) = ", round(prob,3) ))
```

# 

# Question 3

A company is considering whether to do a test on its marketing campaign. Direct marketing costs €2.00 to send. The company sells a product for €50 and 34% of that remains after taking costs and other items into account.\
\
**If they decide to roll it out to the population, what's the probability that we make a mistake, i.e., the true response rate is lower than the breakeven response rate? Use a bootstrap with the same seed and number of draws we did in the R notebook.** 

*Provide your answer with three decimals separated by a dot, not a comma (e.g. 0.123).*

The bootstrap code:

```{r, cache=TRUE}
B <- 10000 
mub <- c() 
set.seed(19103)
for (b in 1:B){
  samp_b = sample.int(nrow(ebeer), replace=TRUE) 
  mub <- c(mub, mean(ebeer$respmail[samp_b])) # RESPONSE RATE
}
```

Probability that True Response Rate is lower than Breakeven Response Rate:

```{r}
B <- 10000
set.seed(19103)
post_draws <- rnorm(B,mean(mub),sd(mub)) # random deviates
prob <- round(sum(post_draws<brk)/B, 3)
```

```{r, echo=FALSE}
cat("The probability of being below the breakeven point is", round(prob, 3))
```

We can plot this:

```{r}
xx <- seq(.11,.14,length=1000)

par(mai=c(.8,.8,.2,.2))
hist(mub, main="", xlab="average response rate",
     col=8, border="grey90", freq=FALSE)
abline(v=brk)
text(x = .115,y= 68, paste("P(p < 0.118) = ", round(prob,3) ))
```

# Question 4

A company wants to calculate the value of testing. They have a margin of 25, a marketing cost of 0.40, and a total customer base of 50000. Their sample is 5000 customers. In the past, campaigns have either been a success, with a response rate of .05, or a failure, with a response rate of 0.001. Historically, 80% of past mailings have been a failure.

**Would the company decide to send without a test?**

Our data:

```{r}
m <- 25
c <- 0.40
N <- 50000
n <- 5000

p_s <- 0.05  # with P(Success)=0.2
p_f <- 0.001 # with P(Failure)=0.8

```

**Option Value (Without Test)**:

```{r}
(0.2)*N*(p_s*m-c)+(0.8)*N*(p_f*m-c)
```

# Question 5

A company wants to calculate the value of testing. They have a margin of 25, a marketing cost of 0.40, and a total customer base of 50000. Their sample is 5000 customers. In the past, campaigns have either been a success, with a response rate of .05, or a failure, with a response rate of 0.001. Historically, 80% of past mailings have been a failure.

**What is the profit if the company tests and it is a failure?**

**Option Value (Failure Case Only)**:

```{r}
n*(p_f*m-c)
```

# Question 6

A company wants to calculate the value of testing. They have a margin of 25, a marketing cost of 0.40, and a total customer base of 50000. Their sample is 5000 customers. In the past, campaigns have either been a success, with a response rate of .05, or a failure, with a response rate of 0.001. Historically, 80% of past mailings have been a failure.

**What's the expected profit with a test?**

**Option Value (Total Expected)**:

```{r}
cat("$",(0.2)*N*(p_s*m-c)+(0.8)*n*(p_f*m-c))
```

# Question 7

A company wants to calculate the value of testing. They have a margin of 25, a marketing cost of 0.40, and a total customer base of 50000. Their sample is 5000 customers. In the past, campaigns have either been a success, with a response rate of .05, or a failure, with a response rate of 0.001.  Historically, 80% of past mailings have been a failure.

**If running a test costs 5000, should the company do it?**

```{r}
(0.2)*N*(p_s*m-c) + (0.8)*(n*(p_f*m-c)) - 5000
```
