---
title: "Quiz 2: RFM Analysis"
author: "Daniel Redel"
date: "2023-01-24"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

library(tidyverse)
library(data.table)
library(kableExtra)
library(gtools)
library(VGAM)
library(readr)
```

```{r, include=FALSE, warning=FALSE}
ebeer_test <- read_csv("C:/Users/danny/OneDrive/Análisis Cuantitativo y Econometría/Marketing Analytics/Customer Analytics/Quizes/2 RFM/ebeer_test.csv")

ebeer_rollout <- read_csv("C:/Users/danny/OneDrive/Análisis Cuantitativo y Econometría/Marketing Analytics/Customer Analytics/Quizes/2 RFM/ebeer_rollout.csv")
```

# Question 1

A company is considering using RFM segments to target its rollout. The mailing under consideration costs €0.50 to send; if customers respond, they spend on average €50, of which €15 is margin.

It conducted a test of customers in the file, [ebeer_test.csv](https://tilburguniversity.instructure.com/courses/10919/files/1948975?wrap=1); the customers not in the test can be found in [ebeer_rollout.csv](https://tilburguniversity.instructure.com/courses/10919/files/1948977?wrap=1). For this analysis, create 3 RFM groups (instead of 5 like we did in class). 

**What is the average recency of the group that is most recent, i.e., the customers who purchased most recently?** 

*Provide your answer with **one** decimal separated by a dot, not a comma (e.g. 0.1).*

```{r}
ebeer_test$Rgroup <- quantcut(ebeer_test$R, q = 3)
```

Alternatively:

```{r}
# Create Quantiles 
ntiles <- function(x, bins) {
  quantiles = seq(from=0, to = 1, length.out=bins+1)
  cut(ecdf(x)(x), breaks = quantiles, labels=F)
}
ebeer_test$Rgroup <- ntiles(ebeer_test$R, bins=3)  
```

```{r, warning=FALSE}
recency_stats <- ebeer_test %>% 
  group_by(Rgroup) %>% 
  summarise(n = n(), 
            mean_R = mean(R), sd_R = sd(R), 
            resp_p = mean(respmail, na.rm=TRUE))
recency_stats %>% 
  kbl() %>%
  kable_styling()
```

```{r, echo=FALSE}
cat("The Average Recency of Group [2:10] is", round(recency_stats$mean_R[1],1) )
```

# Question 2

**What is the response probability of that group?** 

*Provide your answer with **two** decimals separated by a dot, not a comma (e.g. 0.17).*

```{r, echo=FALSE}
cat("The Response Rate of Group [2:10] is", round(recency_stats$resp_p[1],2) )
```

# Question 3

Now do the full RFM analysis. 

**How many RFM groups are there**?

*Provide your answer with **zero** decimals (e.g. 17).*

The Code:

```{r}
ntiles <- function(x, bins) {
  quantiles = seq(from=0, to = 1, length.out=bins+1)
  cut(ecdf(x)(x),breaks=quantiles, labels=F)
}

ebeer_test$Rgroup <- ntiles(ebeer_test$R, bins=3)  


dt = data.table(ebeer_test)
nbins = 3
dt[, RFgroup := paste0(as.character(Rgroup), as.character(ntiles(F, bins = nbins))), by = c('Rgroup')]
dt[, RFMgroup := paste0(as.character(RFgroup), as.character(ntiles(M, bins = nbins))), by = c('RFgroup')]

# put it back to data.frame
ebeer_test = data.frame(dt)

# change it to a factor variable
ebeer_test$RFMgroup<-as.factor(ebeer_test$RFMgroup)
```

**Number of RFM Groups:**

```{r}
length(unique(ebeer_test$RFMgroup))
```

# Question 4

**How many of these segments are profitable to send to?** 

*Provide your answer with **zero** decimals (e.g. 12).*

Let's calculate the response rates for each segment:

```{r, warning=FALSE}
respRFM <- ebeer_test %>% 
  group_by(RFMgroup) %>% 
  summarise(n_resp= sum(respmail, na.rm = TRUE), 
            n_mail= sum(mailing, na.rm = TRUE)) %>% 
  mutate(resp_rate = n_resp/n_mail) %>% 
  arrange(desc(resp_rate))
head(respRFM) %>% 
  kbl() %>%
  kable_styling()
```

**Breakeven Point**:

```{r}
c <- 0.5
m <- 15
brk <- c/m
```

[**How many to target?**]{.underline}:

```{r}
sum(respRFM$resp_rate >= brk)
```

```{r}
respRFM <- as.data.frame(respRFM)

bp <- barplot(respRFM[,4], 
              main="response by RFM group", 
              xlab="RFM Group", ylab="average response", xaxt="n")
axis(1, at = bp[,1], labels=respRFM[,1], cex.axis=0.7, las=2)

abline(h=brk)
text(85, brk, "breakeven", cex=1, pos=3, col="black")
```

# Question 5

**What fraction of total variation in responses is explained by the RFM model?** 

*Provide your answer with **three** decimals separated by a dot, not a comma (e.g. 0.127)*

```{r}
RFM_model <- lm(respmail ~ RFMgroup, data = ebeer_test)
round(summary(RFM_model)$r.squared,3)
```

# Question 6

**How many customers in the roll-out sample should be targeted?** 

*Provide your answer with **zero** decimals (e.g. 12).*

Let's consider now the roll-out data. We need first to *create the segments*:

```{r}
ntiles <- function(x, bins) {
  quantiles = seq(from=0, to = 1, length.out=bins+1)
  cut(ecdf(x)(x),breaks=quantiles, labels=F)
}

ebeer_rollout$Rgroup <- ntiles(ebeer_rollout$R, bins=3)  


dt = data.table(ebeer_rollout)
nbins = 3
dt[, RFgroup := paste0(as.character(Rgroup), as.character(ntiles(F, bins = nbins))), by = c('Rgroup')]
dt[, RFMgroup := paste0(as.character(RFgroup), as.character(ntiles(M, bins = nbins))), by = c('RFgroup')]

# put it back to data.frame
ebeer_rollout = data.frame(dt)

# change it to a factor variable
ebeer_rollout$RFMgroup<-as.factor(ebeer_rollout$RFMgroup)
```

We want to *predict their response rates*:

```{r}
ebeer_rollout$RFMpred <- predict(RFM_model, ebeer_rollout)
sum(ebeer_rollout$RFMpred >= brk)
```

```{r}
rollout_respRFM <- ebeer_rollout %>% 
  group_by(RFMgroup) %>% 
  summarise(n_resp= sum(respmail, na.rm = TRUE), 
            n_mail= sum(mailing, na.rm = TRUE)) %>% 
  mutate(resp_rate = n_resp/n_mail) %>% 
  arrange(desc(resp_rate))
```

```{r}
rollout_respRFM <- as.data.frame(rollout_respRFM)

bp <- barplot(rollout_respRFM[,4], 
              main="response by RFM group", 
              xlab="RFM Group", ylab="average response", xaxt="n")
axis(1, at = bp[,1], labels=rollout_respRFM[,1], cex.axis=0.7, las=2)

abline(h=brk)
text(85, brk, "breakeven", cex=1, pos=3, col="black")
```

# Question 7

**What is the return on investment of rolling out to those segments?** 

*Provide your answer with **zero** decimals **without** the percent sign (e.g. 120).*

**Profit per Consumer**:

```{r}
ebeer_rollout <- ebeer_rollout %>% 
	mutate(RFMprofit = case_when(RFMpred >= brk ~ RFMpred*m-c, TRUE ~ 0))
```

**Sum of Profits**:

```{r}
sum_profit <- sum(ebeer_rollout$RFMprofit)
sum_profit
```

**Costs per Consumer**:

```{r}
ebeer_rollout <- ebeer_rollout %>% 
	mutate(RFMcost = case_when(RFMpred >= brk ~ c, TRUE ~ 0))
```

**Sum of Costs**:

```{r}
sum_cost <- sum(ebeer_rollout$RFMcost)
sum_cost 
```

**ROI**:

```{r, echo=FALSE}
cat("The ROI is", round(sum_profit / sum_cost*100, 0), "%")
```

# Question 8

Use a beta-binomial model to shrink the segment estimates in the test data.  

**How many extra segments would you target using this method that you wouldn't otherwise target?**

*Provide your answer with **zero** decimals (e.g. 12).*

First, we need the non-response data:

```{r}
respRFM <- respRFM %>% 
  mutate(n_nonresp = n_mail-n_resp) %>% 
  relocate(n_nonresp, .after=n_resp)
```

Likelihood Function to fit the data and find **prior response rate**:

```{r}
fit <- vglm(cbind(respRFM$n_resp,respRFM$n_nonresp) ~ 1, betabinomialff, trace=TRUE)
a <- Coef(fit)[[1]]
b <- Coef(fit)[[2]]
```

```{r, echo=FALSE}
cat("(a,b)=(",a,",", b, ")")
```

We know can estimate the posterior response rates:

```{r}
post_mean_resp <- (a+respRFM$n_resp)/(a+b+respRFM$n_mail)
respRFM <- cbind(respRFM, post_mean_resp)
```

```{r}
plot(respRFM$resp_rate, xaxt="n",col="red",xlab="RFM segments",ylab="response rate and posterior mean response rate")
points(respRFM$post_mean_resp, col='blue')
legend('topleft',legend=c("estimate response rate", "posterior expected response rate"),col=c("red","blue"), pch=1)
axis(1, at = 1:27, labels=respRFM$RFMgroup, cex.axis=0.7, las=2)
abline(h=brk)
text(25, brk, "breakeven", cex=1, pos=3, col="black")
```

In the end, ***how many extra segments should we target?***

```{r}
sum(respRFM$post_mean_resp >= brk) - sum(respRFM$resp_rate >= brk)
```
