---
title: "Quiz 5: Recomendation Systems"
author: "Daniel Redel"
date: "2023-01-25"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("scipen"= 200, "digits"=3, width = 100)
library(kableExtra)
rm(list=ls())
```

# Question 1

Let's use the [**content-based method**]{.underline} to make some recommendations. The last 5 things I watched on Netflix and their attributes were:

```{r}
row_names <- c("FUNNY", "ROMANTIC", "SUSPENSE", "DARK", "RATING")
col_names <- c("SHARP OBJECTS", "ARRESTED DEV.", "ARBITRAGE", "MARGIN CALL", "BOJACK", "ORPHAN BLACK", "HINTERLAND")

item <- matrix( c(0, 1, 0, 1, 1, 1, 0,
                  1, 1, 0, 0, 1, 0, 0,
                  1, 1, 1, 0, 1, 0, 1,
                  1, 0, 1, 1, 0, 1, 1,
                  4, 3, 4, 5, 3, NA, NA), byrow = TRUE, nrow = 5, ncol = 7,
                dimnames = list(row_names, col_names))
item %>% 
  kbl() %>%
  kable_styling()
```

**How similar are my preferences to Orphan Black?** [***R: 0.36***]{.underline}

Calculate the cosine similarity and report to 2 decimal points (e.g. 0.12)

First, we normalize the ratings to the mean:

```{r}
rating <- matrix(c(4, 3, 4, 5, 3), nrow=1,ncol=5)
rating_m <- rating-mean(rating)
rating_m %>% 
  kbl() %>%
  kable_styling()
```

We now estimate how much the ratings changes with attributes:

```{r}
user <- item[1:4,1:5] %*% t(rating_m)/rowSums(item[1:4,1:5])
user %>% 
  kbl() %>%
  kable_styling()
```

Finally, we take the similarity with respect to [**Orphan Black**]{.underline}:

```{r}
new_item <- item[1:4,6:7]

CS <- t(new_item) %*% user / sqrt( colSums(new_item^2) )*sqrt( sum(user^2) )
round(CS,2) %>% 
  kbl() %>%
  kable_styling()
```

# Question 2

**How similar are my preferences to Hinterland?** [***R: 0.21***]{.underline}

*Calculate the cosine similarity and report to 2 decimal points (e.g. 0.12).*

```{r}
CS <- t(new_item) %*% user / sqrt( colSums(new_item^2) )*sqrt( sum(user^2) )
CS %>% 
  kbl() %>%
  kable_styling()
```

# Question 3

**What should Netflix recommend to me, based on the content-based method?**

```{r, echo=FALSE}
cat("Orphan Black, because is closer to 1")
```

# Question 4

Now, let's do the same exercise through the lens of collaborative filtering (on users, not items).

```{r}
row_names<-c("George", "Adam", "Ben", "Cam", "Dan")
col_names<-c("Sharp Obj", "Arrested Dev", "Arbitrage", "Margin C", "Bojack", "Orphan B", "Hinterland")

util <- matrix(c(4,3,4,5,3,NA,NA,
                 4,3,4,4,3,NA,NA,
                 3,4,3,1,3,5,NA,
                 4,4,4,4,4,2,4,
                 2,1,2,3,1,NA,3),
               byrow = TRUE, nrow = 5, ncol = 7, 
               dimnames=list(row_names,col_names))
util %>% 
  kbl() %>%
  kable_styling()
```

Imagine we want to again make a recommendation as to which movie George should see, based on his ratings and those of **Adam**, **Ben**, **Cam**, and **Dan**:

**Who is most useful in predicting George's rating for Orphan Black?**

We take the relevant users:

```{r}
m <- cor( t(util), use="pairwise.complete.obs")

# The relevant row
users_corr <- m[row=c("Adam","Ben","Cam", "Dan"), col=c("Adam")]
users_corr %>% 
  kbl() %>%
  kable_styling()

```

We normalize the ratings:

```{r}
util_n <- util - rowMeans(util, na.rm=TRUE) #normalize

# The Movies we care:
movies_corr <- util_n[row=c("Adam","Ben","Cam", "Dan"), col=c("Orphan B", "Hinterland")]
movies_corr %>% 
  kbl() %>%
  kable_styling()

```

```{r, echo=FALSE}
cat("The most useful user for Orphan Black is Ben")
```

# Question 5

**Who is most useful in predicting George's rating for Hinterland?**

```{r}
movies_corr %>% 
  kbl() %>%
  kable_styling()
```

```{r, echo=FALSE}
cat("The most useful user for Hinterland is Dan")
```

# Question 6

**According to the collaborative filtering method, should Netflix better recommend Orphan Black or Hinterland?**

Finally, we predict:

```{r}
predm <- users_corr * movies_corr
predm %>% 
  kbl() %>%
  kable_styling()
# Take the average
pred <- colMeans(predm, na.rm=TRUE)
pred %>% 
  kbl() %>%
  kable_styling()

```

```{r}
predm<-m[row=c("Ben","Dan"),col=c("George")]*util_n[row=c("Ben","Dan"),col=c("Orphan B", "Hinterland")]
predm %>% 
  kbl() %>%
  kable_styling()
# Take the average
pred <- colMeans(predm, na.rm=TRUE)
pred %>% 
  kbl() %>%
  kable_styling()
```
