---
title: "Assignment 6: CLV - Contractual Settings"
author: "Daniel Redel"
date: "2022-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(knitr)
library(kableExtra)
library(Metrics)
options("scipen"=100, "digits"=3, width = 150)
```

# Question 1:

Netflix charges 13 euros per month. Variable costs including billing, server capacity, and marketing spending are 2 euros per month. The monthly discount rate is .01. The churn rate is 0.009 per month.

**What is the residual lifetime value immediately after renewal (the next payment is due the following month)?**

*Provide your answer with zero decimals (e.g. 120 or -120)*.

```{r}
m <- 13 - 2
d <- 0.01
p <- 1 - 0.009  # churning is the opposite of retention
```

E(RLV) just the time **after** the next renewal: $$
E(RLV)=\frac{m(1+d)}{1+d-p}\times \frac{p}{(1+d)}
$$

```{r}
RLV_geo <- m*(1+d)*p/( (1+d-p)*(1+d) ) 
RLV_geo
```

# Question 2:

Blue Apron is a meal box service, similar to HelloFresh and MarleySpoon. subscribers order several meals a month from the service which are delivered to their homes. Subscribers pay 40 euros a month, where 15 goes to variable costs. The monthly discount rate is .01. Starting with a representative 100 customers in period 0, the number that survives over the following 6 months are given below.

| **Months since Joining** | **Number of Subscribers** |
|--------------------------|---------------------------|
| 0                        | 100                       |
| 1                        | 67                        |
| 2                        | 50                        |
| 3                        | 41                        |
| 4                        | 34                        |
| 5                        | 31                        |
| 6                        | 28                        |

**How much higher is the actual retention rate at 6 months than at 1 month?**

*Provide your answer with two decimals separated by a dot, not a comma (e.g. 0.12)*.

First, we create the dataset:

```{r}
active_subs=c(100,
              67,
              50,
              41,
              34,
              31,
              28)
data <- cbind(0:6, active_subs)
colnames(data) <- c("Period", "Active Subscribers")
data %>% 
  kbl() %>%
  kable_styling()
```

$$
r(t) = P(T>t \mid T>t-1) = \frac{S(t)}{S(t-1)} 
$$ Survivor Function:

```{r}
lost <- -diff(active_subs)

S <- active_subs[1:11]/active_subs[1] # Survivor Function

data <- cbind(data, S)
```

Retention Rate:

```{r}

r <- S[2:11]/S[1:10] # DEFINITION

data <- cbind(data, c(NA,r))
colnames(data)[[4]] <- "r"
data %>% 
  kbl() %>%
  kable_styling()
```

**How much higher is the actual retention rate at 6 months than at 1 month?** Answer:

```{r}
round(0.903-0.670,2)
```

Retention Rate Plot:

```{r}
par(mai=c(.9,.8,.2,.2))
plot(0:10, S, type="b",ylab="Survival Function S(t)", xlab="Period",ylim=par("yaxp")[1:2]) 

par(mai=c(.9,.8,.2,.2))
plot(1:10, r, type="b",ylab="retention rate r(t)", xlab="Period",ylim=par("yaxp")[1:2]) 
```

# Question 3:

Fit the Beta-Geometric model to the retention data using only data from the beginning until (and including) 4 months since joining. For your estimation use starting values a = 1 and b = 1.

**What are the estimates of a and b?**

*Provide your answer with three decimals separated by a dot, not a comma (e.g. 0.123)*.

```{r}
r_sBG=function(a,b,t){
    (b+t-1)/(a+b+t-1)
}

lost <- -diff(active_subs[1:5])
active <- active_subs[1:5][-1]
```

```{r}
loop.lik <- function(params) {
a <- params[1]
b <- params[2]

ll <- 0
for (i in 1:length(lost)) {
    ll <- ll+lost[i]*log(beta(a+1,b+i-1)/beta(a,b))
}
ll <- ll+active[i]*log(beta(a,b+i)/beta(a,b))
return(-ll)    #return the negative of the function to maximize likelihood
} 

#find parameters for a and b with optim
sBG <- optim(par=c(1,1),loop.lik)

a<-sBG$par[1]
b<-sBG$par[2]

a
b
```

```{r}
#calculate retention using model parameters
t <- 1:length(active)
r_pred <- r_sBG(a,b,t)
S_pred <- c(1,cumprod(r_pred)) # predicted survivor function
```

```{r}
r <- S[2:11]/S[1:10] # DEFINITION

# plot actual and predicted retention rate
par(mfrow=c(1,1))
plot(t,r_pred,ylab="retention rate",xlab="period",type="b", xaxt="none", ylim = c(.55,1))
lines(t, r[1:4], type="b", col="blue")
axis(1, seq(0,10,1))
legend('right',legend=c("sBG", "actual"),col=c("black","blue"), pch=c(1,1))

```

# Question 4:

Our holdout sample is everything after period 4. **For the model fit in Q3, calculate the mean absolute error of the holdout retention rate**.

*Provide your answer with two decimals separated by a dot, not a comma (e.g. 0.12)*.

First, we predict the retention rate in the holdout sample:

```{r}
r_sBG=function(a,b,t){
    (b+t-1)/(a+b+t-1)
}

t <- 1:length(active_subs[-1])
r_pred <- r_sBG(a,b,t)
```

We now evaluate using MAE Criteria:

```{r}
round(mae(r[5:6], r_pred[5:6]),2)

#Dataset
data.fit <- cbind(pred=r_pred, actual=r[1:6])
data.fit %>% 
  kbl() %>%
  kable_styling()

```

```{r}
par(mfrow=c(1,1))
plot(t,r_pred,ylab="retention rate",xlab="period",type="b", xaxt="none", ylim = c(.55,1))
lines(t, r[1:6], type="b", col="blue")
axis(1, seq(0,10,1))
legend('right',legend=c("sBG", "actual"),col=c("black","blue"), pch=c(1,1))
```

# Question 5:

According to your model fit in Q3, **what percentage of customers will be active one year after the first signup?**

*Provide your answer with two decimals without the percent sign separated by a dot, not a comma (e.g. 0.12)*.

```{r}
t <- 1:12
r_pred <- r_sBG(a,b,t)

S_pred <- c(1,cumprod(r_pred)) 
S_pred[13]

cbind(retention=c(1,r_pred), survivor=S_pred) %>% 
  kbl() %>%
  kable_styling()
```

# Question 6:

**Up to how much would Blue Apron spend to rescue a customer who has renewed once, immediately after that first renewal?**

Use 200 time periods for your simulation and round to whole euros.

*Provide your answer with zero decimals (e.g. 120 or -120)*.

Let's try this out for someone who has renewed $\tau=1$ times:

```{r}
tau <- 1
m <- 40 - 15
d <- 0.01
t <- seq(1,tau+200)

r_pred <- r_sBG(a,b,t)
S_pred <- cumprod(r_pred)

S_shift <- S_pred[(tau+1):length(S_pred)]  # survival function from tau + 1 until T

dis <- 1/(1+d)^(t(1:200)-1) # discount rate

RLV_sBG <- sum(m*S_shift/S_pred[tau]*dis)  # sum of margin x S(tau + t)/ S(tau) x discount
RLV_sBG
```

# Question 7:

**How much more is that amount (in Q6) than the customer lifetime value for a new customer?**

*Provide your answer with zero decimals (e.g. 120 or -120)*.

```{r}
t <- seq(1,200) # time periods
r_pred <- r_sBG(a,b,t) # predicted retention rate
S_pred <- c(1,cumprod(r_pred)[1:199]) # predicted survivor function

dis <- 1/(1+d)^(t-1) # discount factor, first term is present so no discounting

CLV_sBG <- sum(m*S_pred*dis) # the sum of margin x survivor x discount factor

RLV_sBG - CLV_sBG 
```
