---
title: "Quiz 6: CLV - Contractual Settings"
author: "Daniel Redel"
date: "2023-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(knitr)
library(kableExtra)
library(Metrics)
options("scipen"=100, "digits"=3, width = 150)
```

# Question 1

Consider Netflix. A month-long subscription from a customer generates a profit flow of 10 per month, and they use a 0.008 discount rate. Below is a cohort of subscribers who all began their subscription in January 2020: 1000 initially joined; a month later 631 of them did not cancel, etc.

```{r}
active_cust=c(1000,
              631,
              468,
              382,
              326,
              289,
              262,
              241,
              223,
              207,
              194)

cbind(0:10,active_cust) %>% 
  kbl() %>%
  kable_styling()
```

**What is the actual retention rate and actual survival function at [10 months]{.underline} since joining?**

*Provide your answers with a dot and three decimals (e.g. 0.234)*

**Survivor Function**:

```{r}
S <- active_cust/active_cust[1]
S
```

**Retention Rate**:

```{r}
r <- S[2:11]/S[1:10]
r
```

**Final Data**:

```{r}
cbind(0:10, S, c(NA,r)) %>% 
  kbl() %>%
  kable_styling()
```

# Question 2

Fit the Beta-Geometric model to the retention data using only data from the beginning until (and including) 4 months since joining. For your estimation use starting values a = 1 and b = 1.

**What are the estimates of a and b?**  

*Provide your answers with a dot and three decimals (e.g. 0.123)*

[**Calibration Sample**]{.underline}:

```{r}
lost <- -diff(active_cust[1:5])
active <- active_cust[1:5][-1]
```

[**Likelihood Estimation**]{.underline}:

```{r}
loop.lik <- function(params) {
a <- params[1]
b <- params[2]

ll <- 0
for (i in 1:length(lost)) {
    ll <- ll+lost[i]*log(beta(a+1,b+i-1)/beta(a,b))
}
ll <- ll+active[i]*log(beta(a,b+i)/beta(a,b))
return(-ll)  #return the negative of the function to max. LL
} 

#find parameters for a and b with optim
sBG <- optim(par=c(1,1),loop.lik)
```

```{r}
a <- sBG$par[1]
b <- sBG$par[2]

```

```{r, echo=FALSE}
cat("a=", round(a,3), "b=", round(b,3))
```

# Question 3

Our holdout sample is everything after period 4.

**How well does the model fit the data from period 5 onward?**  Make a graph. Use the model estimates to predict retention and survival function at 10 months since joining.

*Provide your answers with a dot and three decimals (e.g. 0.234)*

[**Retention Rate (Prediction)**]{.underline}:

```{r}
r_sBG = function(a,b,t){
    (b+t-1)/(a+b+t-1)
}
# Prediction
t <- seq(1:10)
r_pred <- r_sBG(a,b,t)
```

```{r}
par(mfrow=c(1,1))
plot(t, r_pred,ylab="retention rate",xlab="period",type="b", xaxt="none", ylim = c(.55,1))
lines(t, r, type="b", col="blue")
axis(1, seq(0,10,1))
legend('right',legend=c("sBG", "actual"),col=c("black","blue"), pch=c(1,1))
```

[**Survivor Function (Prediction)**]{.underline}:

```{r}
S_pred <- c(1,cumprod(r_pred)) 
S_pred
```

```{r}
par(mfrow=c(1,1))
plot(seq(0:10), S_pred, ylab="retention rate",xlab="period",type="b", xaxt="none", ylim = c(0,1))
lines(seq(0:10),S, type="b", col="blue")
axis(1, seq(0,10,1))
legend('right',legend=c("sBG", "geo", "actual"),col=c("black","red","blue"), pch=c(1,1))
```

# Question 4

Up to how much would Netflix pay to acquire a ***new customer*** with a value according to that implied by the model? Use 200 months in your calculation.

**Round to the nearest whole number**.

CLV:

```{r}
d <- 0.008
m <- 10
t <- seq(1,200) # time periods
r_pred <- r_sBG(a,b,t) # predicted retention rate
S_pred <- c(1,cumprod(r_pred)[1:199])

dis <- 1/(1+d)^(t-1) # discount factor, first term is present so no discounting

CLV_sBG <- sum(m*S_pred*dis) # the sum of margin x survivor x discount factor
CLV_sBG
```

# Question 5

**Now calculate CLV using the geometric model that assumes a constant retention rate.**

To estimate that constant rate, you can use the results of the BG model. The average retention rate in the population at time 0 is b/(a+b). Use that quantity as your estimate of the retention rate for the geometric model. 

*Give the whole number.*

```{r}
geoCLV<-function(p,m,d){
  m*(1+d)/(1+d-p)
}

p <- b/(a+b) # its B! not A
geo <- geoCLV(p, m, d) # use that estimate
round(geo,0)
```

# Question 6

Back to the original BG coefficients.  

**What is the expected residual lifetime value of a customer who has renewed once, standing just before time 2, when he or she makes his or her second renewal decision?** 

*Give the whole number.*

[**RLV**]{.underline}:

```{r}
tau<-1

t<-seq(1,tau+200)
r_pred<-r_sBG(a,b,t)
S_pred<-cumprod(r_pred)
```

```{r}
S_shift <- S_pred[(tau+1):length(S_pred)]  # survival function from tau + 1 until T
dis <- 1/(1+d)^(t(1:200)-1) # discount rate

RLV_sBG<-sum(m*S_shift/S_pred[tau]*dis)  
RLV_sBG
```
